Using pretense behavior to explore counterfactual self-simulation

### 2023-01-18 docs/cogsci/PNTKN.Rmd
fixed naming of E2.click_log_with_boards (it was unclear how this structure was created)

author-information: > 
    \author{{\large \bf Matan Mazor (mtnmzor@gmail.com)} \\ Birkbeck, University of london \\ London, WC1E 7HX UK
    \AND {\large \bf Chaz Firestone (chaz@jhu.edu)}
    \AND {\large \bf Ian Phillips (ianphillips@jhu.edu)} \\ Johns Hopkins University \\ Baltimore, MD 21218, USA}

@@

Moreover, deception often involves pretense behavior, which in turn relies on an ability to simulate and mimic one's own behavior under a hypothetical belief state. For example, in order to successfully deceive your friends into thinking that you were surprised by the birthday party they threw for you, it is not sufficient that you are able to reason about their mental states ("I know that they are planning a surprise party, but they don't know that I know that.") — you also need to convincingly simulate and mimic your hypothetical behavior had you not known about the party ("Where would I look first? What would I say? How long would it take me to recover from the surprise?"). This reliance of pretense behavior on self-simulation makes it an ideal opportunity to examine metacognitive knowledge about one's own mental states, and the potential reliance of this knowledge on a self-simulation. By comparing non-pretend and pretend behavior, we can ask which aspects of their own cognitive processes subjects can and cannot effectively simulate, and which aspects are not well-represented in their mental models of their own cognition.

@@

The research complied with all relevant ethical regulations and was approved by the Research Ethics Committee of Johns Hopkins University. 500 Participants were recruited via Prolific (prolific.co) and gave their informed consent prior to their participation. They were selected based on their acceptance rate (>95%) and for being native English speakers. The entire experiment took approximately 20 minutes to complete. Participants' pay was equivalent to an hourly wage of 9.50 USD, in addition to a bonus payment (`r E2.bonus$bonus%>%min()` - `r E2.bonus$bonus%>%max()` USD, mean = `r E2.bonus$bonus%>%mean()%>%printnum()`).

@@

> "In the game Battleships, your task is to sink all ships located in a grid with as few clicks as possible.
What makes the game difficult is that you can’t see the ships; all you can see is a grid of squares, and you have to guess where the ships are. To sink a ship, you need to click on all of the squares it is located in. If you hit part of a ship, the square will turn red. If there is no ship in the square, it will turn blue."

@@

> "This time your goal is different. In this round, we’re going to tell you where the ships are, but **we want you to act like you don’t know this information**. We’ve marked the ships’ locations with a cross, so you’ll know where they are the whole time; but your job is to play the game as if these hints aren’t there. To see how good you are at this, we're going to compare your games to the games of people who actually had no hints, and see how similar they are. We will measure where and when you clicked; if your clicks look similar to people who played like normal (trying to reveal all ships with as few clicks as possible, but without any hints), you'll get bonus points. But if your games look different, you won't get these bonus points. Your number of clicks in this part will not affect your bonus. Only your ability to play like you had no hints."

@@

Pretend games were similar to non pretend games in the total number of clicks — but were they also similar in *where* participants clicked? More specifically, did cell selections in pretend games make sense given the limited information those participants pretended to have? To ask this, we approximated optimal behavior by calculating the probability that a ship is hidden in each cell given available information, $p(ship(x_{i}))$, and the posterior probability that one should click on a square, assuming a uniform prior over cells $P(x_{i})=\frac{p(Ship(x_i))}{\sum_{j=1}^{k}p(Ship(x_j))}$. Critically, in modeling pretend games we did not treat hints as part of this available information for extracting $p(ship(x_{i}))$, because an optimal player should ignore hints in choosing where to click next. Given this posterior map, a rational player should choose cells where $P(x_{i})$ is high [this behavior is not strictly optimal, but approximates optimal behavior in most cases; @audinot2014optimal, Section 3.3]. To quantify optimality, before each cell selection we computed the posterior probability map for all 'unknown' cells. Then, we ranked cells from high to low according to their posterior probability and recorded the rank of the chosen cell: a lower rank indicating more optimal behavior.

@@

@@

# E2.click_log <-E2.click_log %>%
#   dplyr::select(subj_id,genuine_first,test_part,grid_number,click_number,i,j,RT,hit_bin) %>% 
#   mutate(flat_position = 1+i*5+j) %>%
#   group_by(subj_id,genuine_first,test_part,grid_number) %>%
#   arrange(click_number) %>%
#   summarise(click_number = click_number,
#             board_state=create_board_states(flat_position,hit_bin),
#             flat_position=flat_position,
#             i=i,
#             j=j,
#             RT=RT,
#             hit_bin=hit_bin) %>%
#   rowwise()%>%
#   mutate(likelihood = get_likelihood(board_state),
#          posterior = get_posterior(likelihood),
#          p_click = get_p_click(posterior,flat_position),
#          entropy = get_posterior_entropy(posterior),
#          p_click_rank = get_p_click_rank(posterior,flat_position));

@@

The previous analyses reveal robust associations between game state and decision latency, that are highly similar, and even exaggerated, in pretend compared to non-pretend games. To get at subtler dynamics in a more direct way, in a follow-up exploratory analysis we focused on decision latencies following a hit. We categorized hits into four types: first hit on a ship, second hit on a ship when the size-three submarine hasn’t been sunk yet, second hit on a ship when the size-three submarine has already been sunk, and third hit on a submarine (see Fig. 3C). In the first case, players know the ship must continue in one of the neighboring cells. In the second case, there is a good chance the ship continues (if this ship turns out to be a submarine). In the third and fourth cases, it is clear that the ship is fully sunk.

@@

Our optimality analysis showed that pretenders’ click selections closely resemble those of non-pretenders, at least in that they are not random, but rather guided by where a ship might be. However,  due to the high number of possible board configurations, data from full games provide limited opportunity to compare cell selections for specific game states. In addition to asking, “What guides cell selections in pretend and non-pretend games?”, we also wanted to ask, “Where exactly would pretenders and non-pretenders click, given a specific board configuration?”.

@@

<!-- Furthermore, compared to non-pretend games, in pretend games participants were more likely to make irrational clicks on cells where a ship could not in fact be hiding (a median of `r E2.num_irrational_clicks%>%filter(test_part=='pretend')%>%pull(nic)%>%median()` per game in pretend versus `r E2.num_irrational_clicks%>%filter(test_part=='nonpretend')%>%pull(nic)%>%median()` in non-pretend games.  `r E2.num_irrational_clicks%>%spread(test_part,nic)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`).  -->
## Judge trials

@@

```
Despite the fact that pretenders were overall successful in mimicking patterns of cell selection and decision latency, they also left many clues for an independent observer to pick up on: they were slower, displayed exaggerated effects of decision uncertainty and hit success on response time, and were less optimal in choosing where to click next. To illustrate, a support vector machine (SVM) algorithm reached an accuracy level of 72% in linearly classifying condition (pretend / non-pretend) based on just three basic summary features: median decision latency, game optimality score, and number of irrational clicks per game.

@@

Many tasks (including writing papers for conferences) require us to overcome the ‘curse of knowledge’ — the challenge of putting ourselves in the shoes of someone who doesn’t know what we know. This is just an example of a more general fact: the ability to simulate counterfactual knowledge states is necessary for efficient and flexible communication. To do this, we must have an internal self-model that can generate thoughts and behavior for hypothetical beliefs, desires, and perceptions.

Research on Bayesian Theory of Mind provides some support for the existence of such a model by measuring subjects' ability to infer beliefs and desires from observed behavior, either explicitly [@baker2009action; @baker2017rational], or implicitly [@onishi2005fifteen; @liu2017ten]. Here we have put forward a complementary approach: Asking participants to generate behavior based on a counterfactual mental state - in this case, a counterfactual knowledge state in which a known piece of information is unknown. Instead of relying on model inversion (e.g., “Which belief states would give rise to this behavior?”), we ask participants to run the model forward, taking beliefs and desires as input and producing behavior as output. Due to the unconstrained space of possible behaviors in our task (cell selections x decision latencies), successfully pretending not to know demands a rich model of cognition, and is much harder to achieve based on a quasi-scientific theory of mental states [@gopnik1994theory].

Our subjects rose to the challenge. Their pretense behavior, although not perfect, closely resembled that of non-pretenders — not only in gross measures such as total number of clicks, but also in subtler patterns of cell selection and decision latency. Our interpretation of this finding is that participants engaged in rich and relatively accurate counterfactual self-simulation: “What would I have done if I didn’t have this piece of knowledge?”

An alternative interpretation is that instead of simulating a counterfactual knowledge state, participants actively suppressed or ignored unwanted knowledge, such that their entire cognitive machinery was available to play the game. This does not require any self-modeling or self-simulation, beyond the knowledge that suppressing knowledge can be beneficial for pretending not to know something. While we cannot fully rule out this interpretation, we think it is unlikely to explain participants’ pretense ability in our data, for two indirect reasons. First, suppressing thoughts on demand is notoriously difficult  [@wegner1987paradoxical], and our first-person experience of performing the task is that suppressing knowledge of ship locations is hardly possible. Second, when asked how they had performed the task in a debrief question, the responses of most participants were aligned with a self-simulation account. In future work, we intend to more directly rule out this knowledge-suppression hypothesis, and well as to generalize our results to other games.

In conclusion, we propose that pretense behavior provides a unique opportunity to directly test the richness and accuracy of subjects’ models of their own cognitive processes, and find evidence for accurate pretending that surprisingly does not translate to an ability to detect pretense. While more work is needed to exclude alternative explanations, we interpret our findings as providing preliminary evidence for a remarkable capacity for counterfactual self-simulation.

@@

\setlength{\parindent}{-0.1in} 
\setlength{\leftskip}{0.125in}
\noindent

>>>>

author-information: > 
    \author{{\large \bf Matan Mazor (mtnmzor@gmail.com)} \\ Birkbeck, University of London \\ London, WC1E 7HX UK
    \AND {\large \bf Chaz Firestone (chaz@jhu.edu)}
    \AND {\large \bf Ian Phillips (ianphillips@jhu.edu)} \\ Johns Hopkins University \\ Baltimore, MD 21218, USA}

@@

Moreover, deception often involves pretense behavior, which in turn relies on an ability to simulate and mimic one's own behavior under a hypothetical belief state. For example, in order to successfully deceive your friends into thinking that you were surprised by the birthday party they threw for you, it is not sufficient that you are able to reason about their mental states ("I know that they are planning a surprise party, but they don't know that I know that.") --- you also need to convincingly simulate and mimic your hypothetical behavior had you not known about the party ("Where would I look first? What would I say? How long would it take me to recover from the surprise?"). This reliance of pretense behavior on self-simulation makes it an ideal opportunity to examine metacognitive knowledge about one's own mental states, and the potential reliance of this knowledge on a self-simulation. By comparing non-pretend and pretend behavior, we can ask which aspects of their own cognitive processes subjects can and cannot effectively simulate, and which aspects are not well-represented in their mental models of their own cognition.

@@

The research complied with all relevant ethical regulations and was approved by the Research Ethics Committee of Johns Hopkins University. 500 Participants were recruited via Prolific (prolific.co) and gave their informed consent prior to their participation. They were selected based on their acceptance rate (\>95%) and for being native English speakers. The entire experiment took approximately 20 minutes to complete. Participants' pay was equivalent to an hourly wage of 9.50 USD, in addition to a bonus payment (`r E2.bonus$bonus%>%min()` - `r E2.bonus$bonus%>%max()` USD, mean = `r E2.bonus$bonus%>%mean()%>%printnum()`).

@@

> "In the game Battleships, your task is to sink all ships located in a grid with as few clicks as possible. What makes the game difficult is that you can't see the ships; all you can see is a grid of squares, and you have to guess where the ships are. To sink a ship, you need to click on all of the squares it is located in. If you hit part of a ship, the square will turn red. If there is no ship in the square, it will turn blue."

@@

> "This time your goal is different. In this round, we're going to tell you where the ships are, but **we want you to act like you don't know this information**. We've marked the ships' locations with a cross, so you'll know where they are the whole time; but your job is to play the game as if these hints aren't there. To see how good you are at this, we're going to compare your games to the games of people who actually had no hints, and see how similar they are. We will measure where and when you clicked; if your clicks look similar to people who played like normal (trying to reveal all ships with as few clicks as possible, but without any hints), you'll get bonus points. But if your games look different, you won't get these bonus points. Your number of clicks in this part will not affect your bonus. Only your ability to play like you had no hints."

@@

Pretend games were similar to non pretend games in the total number of clicks --- but were they also similar in *where* participants clicked? More specifically, did cell selections in pretend games make sense given the limited information those participants pretended to have? To ask this, we approximated optimal behavior by calculating the probability that a ship is hidden in each cell given available information, $p(ship(x_{i}))$, and the posterior probability that one should click on a square, assuming a uniform prior over cells $P(x_{i})=\frac{p(Ship(x_i))}{\sum_{j=1}^{k}p(Ship(x_j))}$. Critically, in modeling pretend games we did not treat hints as part of this available information for extracting $p(ship(x_{i}))$, because an optimal player should ignore hints in choosing where to click next. Given this posterior map, a rational player should choose cells where $P(x_{i})$ is high [this behavior is not strictly optimal, but approximates optimal behavior in most cases; @audinot2014optimal, Section 3.3]. To quantify optimality, before each cell selection we computed the posterior probability map for all 'unknown' cells. Then, we ranked cells from high to low according to their posterior probability and recorded the rank of the chosen cell: a lower rank indicating more optimal behavior.

@@

# a <-E2.click_log %>%
#   dplyr::select(subj_id,genuine_first,test_part,grid_number,click_number,i,j,RT,hit_bin) %>%
#   mutate(flat_position = 1+i*5+j) %>%
#   group_by(subj_id,genuine_first,test_part,grid_number) %>%
#   arrange(click_number) %>%
#   summarise(click_number = click_number,
#             board_state=create_board_states(flat_position,hit_bin),
#             flat_position=flat_position,
#             i=i,
#             j=j,
#             RT=RT,
#             hit_bin=hit_bin) %>%
#   rowwise()%>%
#   mutate(likelihood = get_likelihood(board_state),
#          posterior = get_posterior(likelihood),
#          p_click = get_p_click(posterior,flat_position),
#          entropy = get_posterior_entropy(posterior),
#          p_click_rank = get_p_click_rank(posterior,flat_position));

@@

@@

The previous analyses reveal robust associations between game state and decision latency, that are highly similar, and even exaggerated, in pretend compared to non-pretend games. To get at subtler dynamics in a more direct way, in a follow-up exploratory analysis we focused on decision latencies following a hit. We categorized hits into four types: first hit on a ship, second hit on a ship when the size-three submarine hasn't been sunk yet, second hit on a ship when the size-three submarine has already been sunk, and third hit on a submarine (see Fig. 3C). In the first case, players know the ship must continue in one of the neighboring cells. In the second case, there is a good chance the ship continues (if this ship turns out to be a submarine). In the third and fourth cases, it is clear that the ship is fully sunk.

@@

Our optimality analysis showed that pretenders' click selections closely resemble those of non-pretenders, at least in that they are not random, but rather guided by where a ship might be. However, due to the high number of possible board configurations, data from full games provide limited opportunity to compare cell selections for specific game states. In addition to asking, "What guides cell selections in pretend and non-pretend games?", we also wanted to ask, "Where exactly would pretenders and non-pretenders click, given a specific board configuration?".

@@

<!-- Furthermore, compared to non-pretend games, in pretend games participants were more likely to make irrational clicks on cells where a ship could not in fact be hiding (a median of `r E2.num_irrational_clicks%>%filter(test_part=='pretend')%>%pull(nic)%>%median()` per game in pretend versus `r E2.num_irrational_clicks%>%filter(test_part=='nonpretend')%>%pull(nic)%>%median()` in non-pretend games.  `r E2.num_irrational_clicks%>%spread(test_part,nic)%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`).  -->

## Judge trials

@@

Despite the fact that pretenders were overall successful in mimicking patterns of cell selection and decision latency, they also left many clues for an independent observer to pick up on: they were slower, displayed exaggerated effects of decision uncertainty and hit success on response time, and were less optimal in choosing where to click next. To illustrate, a support vector machine (SVM) algorithm reached an accuracy level of 72% in linearly classifying condition (pretend / non-pretend) based on just three basic summary features: median decision latency, game optimality score, and number of irrational clicks per game.

@@

Many tasks (including writing papers for conferences) require us to overcome the 'curse of knowledge' --- the challenge of putting ourselves in the shoes of someone who doesn't know what we know. This is just an example of a more general fact: the ability to simulate counterfactual knowledge states is necessary for efficient and flexible communication. To do this, we must have an internal self-model that can generate thoughts and behavior for hypothetical beliefs, desires, and perceptions.

Research on Bayesian Theory of Mind provides some support for the existence of such a model by measuring subjects' ability to infer beliefs and desires from observed behavior, either explicitly [@baker2009action; @baker2017rational], or implicitly [@onishi2005fifteen; @liu2017ten]. Here we have put forward a complementary approach: Asking participants to generate behavior based on a counterfactual mental state - in this case, a counterfactual knowledge state in which a known piece of information is unknown. Instead of relying on model inversion (e.g., "Which belief states would give rise to this behavior?"), we ask participants to run the model forward, taking beliefs and desires as input and producing behavior as output. Due to the unconstrained space of possible behaviors in our task (cell selections x decision latencies), successfully pretending not to know demands a rich model of cognition, and is much harder to achieve based on a quasi-scientific theory of mental states [@gopnik1994theory].

Our subjects rose to the challenge. Their pretense behavior, although not perfect, closely resembled that of non-pretenders --- not only in gross measures such as total number of clicks, but also in subtler patterns of cell selection and decision latency. Our interpretation of this finding is that participants engaged in rich and relatively accurate counterfactual self-simulation: "What would I have done if I didn't have this piece of knowledge?"

An alternative interpretation is that instead of simulating a counterfactual knowledge state, participants actively suppressed or ignored unwanted knowledge, such that their entire cognitive machinery was available to play the game. This does not require any self-modeling or self-simulation, beyond the knowledge that suppressing knowledge can be beneficial for pretending not to know something. While we cannot fully rule out this interpretation, we think it is unlikely to explain participants' pretense ability in our data, for two indirect reasons. First, suppressing thoughts on demand is notoriously difficult [@wegner1987paradoxical], and our first-person experience of performing the task is that suppressing knowledge of ship locations is hardly possible. Second, when asked how they had performed the task in a debrief question, the responses of most participants were aligned with a self-simulation account. In future work, we intend to more directly rule out this knowledge-suppression hypothesis, and well as to generalize our results to other games.

In conclusion, we propose that pretense behavior provides a unique opportunity to directly test the richness and accuracy of subjects' models of their own cognitive processes, and find evidence for accurate pretending that surprisingly does not translate to an ability to detect pretense. While more work is needed to exclude alternative explanations, we interpret our findings as providing preliminary evidence for a remarkable capacity for counterfactual self-simulation.

@@

```{=tex}
\setlength{\parindent}{-0.1in} 
\setlength{\leftskip}{0.125in}
```
\noindent
